---
title: "Get Started with biopalette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with biopalette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 3,
  fig.align  = "center"
)
```

## Overview

**biopalette** provides color palettes for biomedical visualization, each sourced from a real image. This guide covers the core workflow: browsing, retrieving, previewing, and using palettes in plots.

## Installation

```{r install, eval = FALSE}
pak::pkg_install("evanbio/biopalette")
```

## Browse Available Palettes

```{r list}
library(biopalette)

list_palettes()
```

`list_palettes()` returns a data frame of all available palettes with their name, type, and number of colors.

`palette_gallery()` renders a visual overview and returns one ggplot per page, so you can display the page you want:

```{r gallery, fig.height = 4}
pages <- palette_gallery(verbose = FALSE)
names(pages)

pages[["qualitative_page1"]]
```

## Retrieve Colors

```{r get}
# Full palette
get_palette("three_body")

# First n colors
get_palette("babel", n = 5)

# Specify type for disambiguation
get_palette("walter_white", type = "diverging")
```

The returned value is a character vector of HEX codes, ready to pass to any plotting function.

## Preview a Palette

```{r preview-rect, fig.height = 2.2}
preview_palette("gene_red", plot_type = "rect")
```

```{r preview-babel, fig.height = 2.2}
preview_palette("babel", plot_type = "rect")
```

```{r preview-circle, fig.height = 2.5}
preview_palette("three_body", plot_type = "circle")
```

## Use in ggplot2

```{r ggplot, fig.height = 4}
library(ggplot2)

ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point(size = 2) +
  scale_color_manual(values = get_palette("three_body")) +
  theme_minimal()
```

For continuous scales, pass the colors to `scale_fill_gradientn()` or `scale_color_gradientn()`. Diverging palettes are the natural fit here:

```{r ggplot-continuous, fig.height = 4}
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  scale_fill_gradientn(colors = get_palette("walter_white")) +
  theme_minimal()
```

## Color Utilities

```{r utils}
hex2rgb("#1688A7")
rgb2hex(c(22, 136, 167))
```

## Function Reference

| Function | Purpose |
|---|---|
| `list_palettes()` | Data frame of all available palettes |
| `palette_gallery()` | Visual gallery of all palettes |
| `get_palette()` | Retrieve colors by name, type, and size |
| `preview_palette()` | Render color swatches |
| `create_palette()` | Add a new palette |
| `compile_palettes()` | Compile all JSONs into a named list |
| `remove_palette()` | Remove a palette by name |
| `hex2rgb()` | HEX to RGB |
| `rgb2hex()` | RGB to HEX |

## Getting Help

- **Documentation**: <https://evanbio.github.io/biopalette/>
- **Issues**: [GitHub Issues](https://github.com/evanbio/biopalette/issues)
- **Function help**: `?get_palette`, `?preview_palette`
