---
title: "Installation Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Reason: this vignette is mostly install/uninstall commands, which must not
# run while the vignette is built. Only the verification chunk is evaluated.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Quick Install

```{r github-install}
# Using pak (recommended)
install.packages("pak")
pak::pkg_install("evanbio/biopalette")

# Using remotes
install.packages("remotes")
remotes::install_github("evanbio/biopalette")
```

**Requires:** R ≥ 4.1.0

---

## Dependencies

All dependencies are installed automatically with the package.

- **cli** — User-facing messages and warnings
- **ggplot2** — Palette preview rendering
- **jsonlite** — JSON read/write for palette storage

---

## Verify Installation

```{r verify, eval = TRUE}
library(biopalette)

packageVersion("biopalette")
list_palettes()
```

---

## Update

```{r update}
pak::pkg_install("evanbio/biopalette")
```

---

## Troubleshooting

### Installation fails on Windows

Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/) for packages that require compilation, then retry.

### Network / Firewall issues

```{r proxy}
Sys.setenv(http_proxy  = "http://your-proxy:port")
Sys.setenv(https_proxy = "https://your-proxy:port")
```

---

## Uninstall

```{r uninstall}
remove.packages("biopalette")
```

---

## Getting Help

- **Documentation**: <https://evanbio.github.io/biopalette/>
- **Issues**: [GitHub Issues](https://github.com/evanbio/biopalette/issues)
