---
title: "Installing biopalette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing biopalette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Requirements

biopalette requires R 4.1 or later. It works on Windows, macOS, and Linux and
does not contain compiled code.

The package uses four runtime dependencies:

- **cli** for user-facing messages;
- **ggplot2** for previews, galleries, and scales;
- **jsonlite** for reading and writing palette collections;
- **scales** for continuous color interpolation.

R installs these dependencies automatically when biopalette is installed.

## Install from GitHub

The development version is available from GitHub. We recommend
[pak](https://pak.r-lib.org/) because it resolves dependencies and reports
installation problems clearly.

```{r install-pak}
install.packages("pak")
pak::pkg_install("evanbio/biopalette")
```

Alternatively, install with remotes:

```{r install-remotes}
install.packages("remotes")
remotes::install_github("evanbio/biopalette")
```

You only need to install `pak` or `remotes` once.

## Verify the installation

Load the package, inspect its version, and list a few bundled palettes:

```{r verify, eval = TRUE}
library(biopalette)

packageVersion("biopalette")
head(list_palettes()[c("name", "type", "n_color")])
```

For a visual check, open the palette gallery in an interactive R session:

```{r verify-gallery}
palette_gallery()
```

## Update

Run the same GitHub installation command to update to the latest development
version:

```{r update}
pak::pkg_install("evanbio/biopalette")
```

Restart R after updating if biopalette was loaded in the current session. This
ensures that R uses the newly installed namespace and package files.

## Troubleshooting

### R cannot install a dependency

Start a fresh R session and retry the installation. If the error identifies a
specific dependency, install that package directly to expose its complete error
message:

```{r dependency-error}
install.packages("packageName")
```

biopalette itself does not require compilation. On Windows, Rtools is needed
only when R must install a dependency from source and that dependency contains
compiled code. Install the version of
[Rtools](https://cran.r-project.org/bin/windows/Rtools/) that matches your R
version if the error explicitly says that build tools are required.

### GitHub cannot be reached

Confirm that the repository is accessible in a web browser and that R can
connect to GitHub. On managed institutional networks, use the proxy or
certificate settings supplied by your system administrator; do not place
credentials in scripts committed to version control.

### R loads an older version

Check the installed version and library location:

```{r old-version}
packageVersion("biopalette")
find.package("biopalette")
.libPaths()
```

Multiple R libraries can contain different copies of the package. Remove the
older copy from the library reported by `find.package()` or install the update
into that library.

## Uninstall

```{r uninstall}
remove.packages("biopalette")
```

If more than one library contains biopalette, pass the relevant library path
through the `lib` argument of `remove.packages()`.

## Getting help

- Read the [package documentation](https://evanbio.github.io/biopalette/).
- Report reproducible problems in [GitHub Issues](https://github.com/evanbio/biopalette/issues).

When reporting an installation problem, include the complete error message and
the output of `sessionInfo()`. Remove tokens, passwords, user names, and other
sensitive paths before posting the output publicly.
