---
title: "Managing custom palette collections"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Managing custom palette collections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3,
  fig.align = "center"
)
```

biopalette includes a curated palette collection, but the same tools can read
and manage a collection owned by a project, laboratory, or organization. A
custom collection is useful when colors carry stable meaning across several
figures or analyses and should be stored, reviewed, and reused like other
project data.

This vignette covers the complete lifecycle of such a collection. For the
bundled palettes and plotting scales, begin with
`vignette("get-started", package = "biopalette")`.

```{r load}
library(biopalette)
```

## Collection structure

A collection is a directory with up to three type subdirectories:

```text
palettes/
├── qualitative/
├── sequential/
└── diverging/
```

Each palette is one JSON file. There is no compiled copy or secondary data
format. A minimal file contains a name, a type, and a vector of HEX colors:

```json
{
  "name": "treatment_groups",
  "type": "qualitative",
  "colors": ["#3E6186", "#EE761F", "#318336"]
}
```

The file name must match `name`, and its parent directory must match `type`.
Palette names are unique across the entire collection, so a name cannot refer
to different palettes under different types.

Choose the type from the meaning of the colors:

- `qualitative` for unordered categories;
- `sequential` for an ordered low-to-high ramp;
- `diverging` for two directions around a meaningful center.

## Create a collection

For this example, create an isolated temporary directory. In a real project,
use a stable path such as `config/palettes` or `assets/palettes` and commit the
JSON files to version control.

```{r collection-directory}
palettes_dir <- tempfile("biopalette-palettes-")
```

`create_palette()` creates the required type directory and writes a validated
JSON file:

```{r create-palettes}
create_palette(
  name = "treatment_groups",
  type = "qualitative",
  colors = c("#3E6186", "#EE761F", "#318336"),
  palettes_dir = palettes_dir
)

create_palette(
  name = "response_blue",
  type = "sequential",
  colors = c("#F2F6FA", "#B9DBF4", "#3A68AE", "#155289"),
  palettes_dir = palettes_dir
)

create_palette(
  name = "effect_balance",
  type = "diverging",
  colors = c("#1991A9", "#A3C5C4", "#E7E9E4", "#D7AD85", "#A65C31"),
  palettes_dir = palettes_dir
)
```

`palettes_dir` is required for every write. This is deliberate: package data
cannot be modified accidentally, and a write always has an explicit owner and
destination.

## Inspect and use the collection

Pass the same directory to any palette-reading function:

```{r inspect-collection}
list_palettes(palettes_dir = palettes_dir)[c("name", "type", "n_color")]
palette_info("response_blue", palettes_dir = palettes_dir)
get_palette("treatment_groups", palettes_dir = palettes_dir)
```

Custom palettes follow exactly the same type-aware `n` rules as bundled ones.
A qualitative palette returns its first `n` category colors, while sequential
and diverging palettes sample the complete ramp:

```{r collection-n}
get_palette("treatment_groups", n = 2, palettes_dir = palettes_dir)
get_palette("response_blue", n = 7, palettes_dir = palettes_dir)
```

They also work directly in ggplot2 scales:

```{r custom-discrete-scale, fig.height = 4}
library(ggplot2)

ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point(size = 2.5) +
  scale_color_biopalette(
    "treatment_groups",
    palettes_dir = palettes_dir
  ) +
  theme_minimal()
```

```{r custom-gradient-scale, fig.height = 4}
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_raster() +
  scale_fill_biopalette_gradient(
    "response_blue",
    palettes_dir = palettes_dir
  ) +
  theme_minimal()
```

The bundled and custom collections remain separate. Omitting `palettes_dir`
always selects the palettes installed with biopalette; supplying it selects
only that custom collection. The two collections are not merged implicitly.

## Update a palette safely

Existing files are protected from accidental replacement:

```{r overwrite-protection, error = TRUE}
create_palette(
  "response_blue",
  "sequential",
  c("#EFF5FB", "#9ECAE1", "#3182BD"),
  palettes_dir = palettes_dir
)
```

Use `overwrite = TRUE` only when replacement is intentional:

```{r overwrite-palette}
create_palette(
  "response_blue",
  "sequential",
  c("#EFF5FB", "#9ECAE1", "#3182BD"),
  palettes_dir = palettes_dir,
  overwrite = TRUE
)

get_palette("response_blue", palettes_dir = palettes_dir)
```

Writes are transactional from the caller's perspective. biopalette writes a
temporary JSON file beside the destination, reads it back through the normal
validator, and commits it only after validation succeeds. If an overwrite
cannot be completed, the previous palette is retained whenever the file system
allows it. The collection cache is invalidated only after a successful write.

## Validation and collaboration

All reading functions use the same collection loader. It checks every JSON file
before returning data and reports the discovered problems together. A malformed
file is not silently skipped, because silently dropping a palette would turn a
data-quality problem into a misleading "not found" error later.

Useful project practices are therefore straightforward:

1. Keep the collection under version control.
2. Review palette type, order, and source meaning along with the HEX values.
3. Use one stable palette name in analysis code instead of copying vectors
   between scripts.
4. Exercise the collection in continuous integration with a read such as:

```{r ci-check, eval = FALSE}
stopifnot(nrow(list_palettes(palettes_dir = "config/palettes")) > 0)
```

This check invokes the same validation path used by plots and palette queries;
it does not create a second validation contract.

## Remove a palette

Deletion also requires an explicit custom collection directory. If `type` is
omitted, biopalette searches all three type directories:

```{r remove-palette}
remove_palette("response_blue", palettes_dir = palettes_dir)
remove_palette(
  "treatment_groups",
  type = "qualitative",
  palettes_dir = palettes_dir
)
```

The function returns `TRUE` invisibly after a successful deletion and `FALSE`
invisibly when the palette is absent. The bundled collection cannot be removed
through this interface because `palettes_dir` has no write default.

```{r remove-final-palette}
remove_palette("effect_balance", palettes_dir = palettes_dir)
unlink(palettes_dir, recursive = TRUE)
```

## Color representation

Palette JSON accepts uppercase or lowercase 6-digit HEX colors and 8-digit HEX
colors with alpha. `hex2rgb()` and `rgb2hex()` provide a lossless conversion
path when colors need to be inspected or exchanged with numeric RGB(A) data:

```{r color-representation}
channels <- hex2rgb(c("#3E6186", "#EE761F80"))
channels
rgb2hex(channels)
```

For color derivation and interpolation, use a color-aware workflow rather than
averaging RGB channels casually. biopalette itself interpolates sequential and
diverging ramps in Lab color space so palette retrieval and ggplot2 gradients
remain consistent.

## Related documentation

- `?create_palette` and `?remove_palette` document the write operations.
- `?get_palette` explains type-aware sampling.
- `?scale_color_biopalette` and `?scale_color_biopalette_gradient` document
  custom scale options.
- [GitHub Issues](https://github.com/evanbio/biopalette/issues) is the place to
  report reproducible collection or validation problems.
