---
title: "Color Palette Management"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Color Palette Management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 2.4,
  fig.align  = "center"
)
```

## Overview

The palette module provides nine functions covering four areas:

| Area | Functions |
|---|---|
| Querying palettes | `get_palette()`, `list_palettes()` |
| Visualization | `preview_palette()`, `palette_gallery()` |
| Managing palettes | `create_palette()`, `remove_palette()`, `compile_palettes()` |
| Color conversion | `hex2rgb()`, `rgb2hex()` |

```{r load}
library(biopalette)
```

Palettes are stored as JSON files organized under three subdirectories —
`sequential/`, `diverging/`, and `qualitative/` — and compiled into the
`palettes` package dataset via `compile_palettes()`.

Every chunk below is evaluated when the vignette is built, so the output you
see is what the current version actually produces.

---

## 1 Querying Palettes

### `get_palette()` — Retrieve a palette by name

Returns a character vector of HEX color codes. If `type` is omitted, the
palette type is auto-detected. Use `n` to take the first N colors from a
larger palette.

```{r get-palette-basic}
# Auto-detect type
get_palette("babel")

# Specify type explicitly, take only the first 5 colors
get_palette("babel", type = "qualitative", n = 5)
```

When `type` is wrong, the error message tells you where the palette actually lives:

```{r get-palette-wrong-type, error = TRUE}
get_palette("walter_white", type = "qualitative")
```

Requesting more colors than the palette contains raises an informative error
rather than silently recycling:

```{r get-palette-too-many, error = TRUE}
get_palette("gene_red", type = "qualitative", n = 5)
```

---

### `list_palettes()` — Browse available palettes

Returns a data frame with columns `name`, `type`, `n_color`, and `colors`.
It reads the **compiled package dataset**, not a directory on disk.

```{r list-palettes-all}
list_palettes()
```

Filter by one or more types with the `type` argument:

```{r list-palettes-filter}
list_palettes(type = "qualitative")

list_palettes(type = c("sequential", "diverging"))
```

Results are sorted by `type`, `n_color`, and `name` by default. Set
`sort = FALSE` to keep the original file order.

---

## 2 Visualization

### `preview_palette()` — Visualize a single palette

Plots a single palette using one of five styles: `"bar"` (default), `"pie"`,
`"point"`, `"rect"`, or `"circle"`. Returns `NULL` invisibly — called for the
plotting side effect.

```{r preview-rect}
preview_palette("gene_red", plot_type = "rect")
```

```{r preview-pie, fig.height = 3.5}
preview_palette("walter_white", type = "diverging", plot_type = "pie")
```

```{r preview-circle, fig.height = 2.5}
preview_palette("three_body", n = 3, plot_type = "circle")
```

Supply `title` to override the default title (which is the palette name):

```{r preview-title}
preview_palette("babel", plot_type = "rect",
                title = "Pan-cancer myeloid cell types")
```

---

### `palette_gallery()` — Browse all palettes at once

Renders a paged gallery of all palettes and returns a named list of ggplot
objects (one per page). Useful for picking colors interactively.

```{r gallery-all}
plots <- palette_gallery()
names(plots)
```

Access individual pages from the returned list:

```{r gallery-access, fig.height = 4}
plots[["qualitative_page1"]]
```

Filter by type and control how many palettes appear per page. Set
`verbose = FALSE` to suppress progress messages when calling
`palette_gallery()` programmatically:

```{r gallery-filter, fig.height = 3}
pages <- palette_gallery(type = "diverging", verbose = FALSE)
pages[["diverging_page1"]]
```

---

## 3 Managing Palettes

### `create_palette()` — Save a custom palette

Writes a named palette as a JSON file under `color_dir/<type>/<name>.json`.
The directory is created automatically if it does not exist.

```{r create-basic}
temp_dir <- file.path(tempdir(), "palettes")

create_palette("my_blues", "sequential", c("#deebf7", "#9ecae1", "#3182bd"),
               color_dir = temp_dir)

create_palette("my_trio", "qualitative",
               c("#E64B35", "#4DBBD5", "#00A087"),
               color_dir = temp_dir)
```

Note that this writes a **file**; it does not register the palette with the
package. `get_palette()` and `list_palettes()` read the compiled dataset, so a
palette in your own directory only becomes visible once that directory is
compiled — see `compile_palettes()` below.

By default, saving over an existing name raises an error. Pass
`overwrite = TRUE` to replace it:

```{r create-overwrite}
create_palette("my_blues", "sequential", c("#c6dbef", "#6baed6", "#2171b5"),
               color_dir = temp_dir, overwrite = TRUE)
```

```{r create-error, error = TRUE}
# Without overwrite = TRUE
create_palette("my_blues", "sequential", c("#deebf7", "#9ecae1", "#3182bd"),
               color_dir = temp_dir)
```

The function invisibly returns a list with `path` and `info` so the result
can be captured when needed.

---

### `remove_palette()` — Delete a palette JSON

Removes a palette JSON file by name. If `type` is omitted, all three type
directories are searched in order.

```{r remove-basic}
remove_palette("my_blues", color_dir = temp_dir)

# Specify type to skip the search
remove_palette("my_trio", type = "qualitative", color_dir = temp_dir)
```

If the palette is not found in any directory, a message is issued and the
function returns `FALSE` invisibly:

```{r remove-missing}
remove_palette("nonexistent", color_dir = temp_dir)
```

---

### `compile_palettes()` — Build the palette dataset

Reads all JSON files under `palettes_dir/sequential/`, `palettes_dir/diverging/`,
and `palettes_dir/qualitative/`, validates them, and returns a structured list.
This is the function used in `data-raw/palettes.R` to rebuild the `palettes`
package dataset via `usethis::use_data()`.

```{r compile}
compiled <- compile_palettes(
  palettes_dir = system.file("extdata", "palettes", package = "biopalette")
)
```

The return value is a named list with three elements — `sequential`,
`diverging`, and `qualitative` — each of which is a named list of HEX vectors:

```{r compile-access}
names(compiled)

compiled$qualitative[["three_body"]]
```

> Validation is strict: a JSON file with a missing field, an unknown type, or
> an invalid HEX code **aborts the whole compile** rather than being skipped,
> so a broken file can never quietly drop a palette out of the dataset.
> Duplicate palette names within the same type are a softer case — they emit a
> message and the last file read wins.

---

## 4 Color Conversion

### `hex2rgb()` — HEX to RGB data frame

Converts a character vector of HEX codes to a data frame with columns
`hex`, `r`, `g`, `b`. Both 6-digit (`#RRGGBB`) and 8-digit (`#RRGGBBAA`)
codes are accepted; the alpha channel is silently ignored.

```{r hex2rgb-basic}
hex2rgb("#1688A7")

hex2rgb(c("#1688A7", "#FF4500", "#339933"))
```

The `#` prefix is required; codes that fail the pattern and `NA` values are
both reported as invalid:

```{r hex2rgb-error-prefix, error = TRUE}
hex2rgb("1688A7")
```

```{r hex2rgb-error-na, error = TRUE}
hex2rgb(c("#1688A7", NA))
```

---

### `rgb2hex()` — RGB to HEX color codes

The symmetric counterpart to `hex2rgb()`. Accepts either a numeric vector of
length 3 or the data frame returned by `hex2rgb()`. Non-integer values are
rounded before conversion.

```{r rgb2hex-vector}
# Single color as a length-3 vector
rgb2hex(c(22, 136, 167))

# Round-trip: HEX -> RGB -> HEX
rgb2hex(hex2rgb(c("#1688A7", "#FF4500")))

# Non-integer values are rounded
rgb2hex(c(21.7, 136.2, 167.4))
```

Typical failure cases for the vector form:

```{r rgb2hex-error-length, error = TRUE}
rgb2hex(c(22, 136))
```

```{r rgb2hex-error-range, error = TRUE}
rgb2hex(c(22, 136, 300))
```

And for the data frame form:

```{r rgb2hex-error-col, error = TRUE}
rgb2hex(data.frame(r = 22, g = 136))
```

```{r rgb2hex-error-df-range, error = TRUE}
rgb2hex(data.frame(r = 22, g = 136, b = 300))
```

---

## 5 A Combined Workflow

The palette and conversion functions compose naturally. The example below picks
a qualitative palette, converts its colors to RGB, lightens them, and saves the
result as a new palette.

```{r workflow-derive}
# 1. Retrieve a qualitative palette
colors <- get_palette("three_body", type = "qualitative")
colors

# 2. Convert to RGB for numeric manipulation
rgb_df <- hex2rgb(colors)
rgb_df

# 3. Lighten each color by blending 50% toward white
rgb_light <- rgb_df
rgb_light[c("r", "g", "b")] <- (rgb_df[c("r", "g", "b")] + 255) / 2

# 4. Convert back to HEX
light_hex <- rgb2hex(rgb_light)
light_hex
```

```{r workflow-save}
# 5. Save the derived palette to your own directory
create_palette("three_body_light", "qualitative", light_hex,
               color_dir = temp_dir)

# 6. Compile that directory to make the palette usable.
#    create_palette() wrote a file; compile_palettes() is what turns a
#    directory of files into the list the query functions work with.
mine <- compile_palettes(temp_dir)
mine$qualitative[["three_body_light"]]
```

```{r workflow-cleanup}
# 7. Clean up
unlink(temp_dir, recursive = TRUE)
```

To make a palette part of the package itself rather than a local directory,
write its JSON under `inst/extdata/palettes/` and rebuild the dataset with
`data-raw/palettes.R`.

---

## Getting Help

- `?get_palette`, `?list_palettes`
- `?preview_palette`, `?palette_gallery`
- `?create_palette`, `?remove_palette`, `?compile_palettes`
- `?hex2rgb`, `?rgb2hex`
- [GitHub Issues](https://github.com/evanbio/biopalette/issues)
